---
title: "Course Project"
output: word_document
editor_options: 
  chunk_output_type: console
---
#Predictive Analytics Course Project Phase 1
##Kellie McLiverty


###Assignment Needs & Data Importation  

Libraries & dataset needed for Assignemnt  
```{r}
options(tidyverse.quiet = TRUE)
library(tidyverse)
library(dplyr)
library(caret)
library(mice) #package for imputation
library(VIM) #visualizing missingness
library(GGally)
library(MASS) #Conflicts with dplyr remember to use dplyr::for select commands
library(leaps)

rain <- read_csv("rain.csv")
str(rain)
```


Factor conversion and recoding  
```{r}
rain = rain%>% mutate(Date =as_factor(as.character(Date)))


rain = rain%>% mutate(RainToday =as_factor(as.character(RainToday)))%>% mutate(RainToday =fct_recode(RainToday,"No" = "0","Yes" = "1"))

rain = rain%>% mutate(RainTomorrow =as_factor(as.character(RainTomorrow)))%>% mutate(RainTomorrow =fct_recode(RainTomorrow,"No" = "0","Yes" = "1"))

rain = rain %>% mutate(WindGustDir = as_factor(WindGustDir)) %>%
mutate(WindGustDir = fct_recode(WindGustDir,
"E" = "1",
"W" = "2",
"N" = "3",
"S" = "4",
"NE" = "5",
"NW" = "6",
"SE" = "7",
"SW" = "8",
"ENE" = "9",
"ESE" = "10",
"WNW" = "11",
"WSW" = "12",
"SSE" = "13",
"SSW" = "14",
"NNE" = "15",
"NNW" = "16"
))

rain = rain %>% mutate(WindDir9am = as_factor(WindDir9am)) %>%
mutate(WindDir9am = fct_recode(WindDir9am,
"E" = "1",
"W" = "2",
"N" = "3",
"S" = "4",
"NE" = "5",
"NW" = "6",
"SE" = "7",
"SW" = "8",
"ENE" = "9",
"ESE" = "10",
"WNW" = "11",
"WSW" = "12",
"SSE" = "13",
"SSW" = "14",
"NNE" = "15",
"NNW" = "16"
))

rain = rain %>% mutate(WindDir3pm = as_factor(WindDir3pm)) %>%
mutate(WindDir3pm = fct_recode(WindDir3pm,
"E" = "1",
"W" = "2",
"N" = "3",
"S" = "4",
"NE" = "5",
"NW" = "6",
"SE" = "7",
"SW" = "8",
"ENE" = "9",
"ESE" = "10",
"WNW" = "11",
"WSW" = "12",
"SSE" = "13",
"SSW" = "14",
"NNE" = "15",
"NNW" = "16"
))


str(rain)
```

###Testing for Missing Data
```{r}
#is.na(rain) Finds all missing data in the dataset commented out for knitting 
vim_plot = aggr(rain, numbers = TRUE, prop = c(TRUE, FALSE),cex.axis=.7) #Plots missing data but prints way too much so it has been commented out for knitting.
```

We have several rows of missing data in this dataset. To have the most accurate data we will need to remove this missing data. Reviewing the vim_plot we can see the columns for Cloud Coverage at 9 am and 3pm are missing loads of data. This could be because the data is not the most important predictor of if it will rain based on if there are clouds at a certain time of day. As such, I will begin by using Column-wise deletion to remove these two columns as well as date as it is not needed.

Column-wise deletion & removing missing data
```{r}
rain = dplyr::select(rain, -Date, -Cloud9am, -Cloud3pm) #creates new dataset without these two cols
vim_plot2 = aggr(rain, numbers = TRUE, prop = c(TRUE, FALSE),cex.axis=.7) #Plots missing data 
```

Aggrigating data to have Average variables
```{r}
rain2<-rain %>% mutate(AvgHumidity = rowMeans(rain[c('Humidity9am', 'Humidity3pm')], na.rm=FALSE))%>% mutate(AvgPressure = rowMeans(rain[c('Pressure9am', 'Pressure3pm')], na.rm=FALSE))%>% mutate(AvgWindSpeed = rowMeans(rain[c('WindSpeed9am', 'WindSpeed3pm')], na.rm=FALSE))%>% mutate(AvgTemp = rowMeans(rain[c('Temp9am', 'Temp3pm')], na.rm=FALSE))


#select only variables relevant to our analysis for final set
rain3<-dplyr::select(rain2, RainTomorrow, RainToday, Rainfall, MaxTemp, MinTemp, WindGustDir, WindGustSpeed, AvgHumidity,AvgPressure,AvgWindSpeed,AvgTemp)

```

Next we will use the "mice" package to do imputation on the rest of the missing data.
```{r}

imp_rain = mice(rain3, m=1, method='pmm', printFlag=FALSE)
summary(imp_rain)
```

Merge the imputed values into our titanic data frame  
```{r}
rain_complete = complete(imp_rain) 
summary(rain_complete)


vim_plot3 = aggr(rain_complete, numbers = TRUE, prop = c(TRUE, FALSE),cex.axis=.7)# Plots missing data .
```

###Looking for important variables
```{r}
ggpairs(rain_complete, columns = c(1:6), cardinality_threshold = 20) #MinTemp to Wind Gust

ggpairs(rain_complete, columns = c(1, 7:11), cardinality_threshold = 20) #

```


###Graphing the Variables in relation to response variable
```{r}
#Looking at the relationship of RainTomorrow to RainToday
ggplot(rain_complete, aes(x=RainToday, fill = RainTomorrow)) + geom_bar() 


#table view
tb = table(rain_complete$RainTomorrow, rain_complete$RainToday) #creates table object
prop.table(tb, margin = 2) #crosstab with proportions
```


```{r}
#Looking at the relationship of RainTomorrow to RainFall **
ggplot(rain_complete, aes(y=Rainfall, x= RainTomorrow)) + geom_point() 

#table view
tb1 = table(rain_complete$RainTomorrow, rain_complete$Rainfall) #creates table object
prop.table(tb1, margin = 2) #crosstab with proportions
```


```{r}
#Looking at the relationship of RainTomorrow to WindGustDir **
ggplot(rain_complete, aes(x= WindGustDir, fill= RainTomorrow)) + geom_bar() 

#table view
tb2 = table(rain_complete$RainTomorrow, rain_complete$WindGustDir) #creates table object
prop.table(tb2, margin = 2) #crosstab with proportions
```


```{r}
#Looking at the relationship of RainTomorrow to WindGustSpeed
ggplot(rain_complete, aes(x= WindGustSpeed, fill= RainTomorrow)) + geom_bar() 

#table view
tb3 = table(rain_complete$RainTomorrow, rain_complete$WindGustSpeed) #creates table object
prop.table(tb3, margin = 2) #crosstab with proportions
```


```{r}
#Looking at the relationship of RainTomorrow to AvgHumidity ***
ggplot(rain_complete, aes(x= AvgHumidity, fill= RainTomorrow)) + geom_bar() 
ggsave("humidity.png")


#table view
tb4 = table(rain_complete$RainTomorrow, rain_complete$AvgHumidity) #creates table object
prop.table(tb4, margin = 2) #crosstab with proportions
```



```{r}
#Looking at the relationship of RainTomorrow to AvgPressure **
ggplot(rain_complete, aes(x=RainTomorrow , y= AvgPressure)) + geom_boxplot() 
ggsave("pressure.png")

#table view
tb6 = table(rain_complete$RainTomorrow, rain_complete$AvgPressure) #creates table object
prop.table(tb6, margin = 2) #crosstab with proportions
```

```{r}
#Looking at the relationship of RainTomorrow to AvgWindSpeed
ggplot(rain_complete, aes(x= AvgWindSpeed, fill= RainTomorrow)) + geom_bar() 

#table view
tb7 = table(rain_complete$RainTomorrow, rain_complete$AvgWindSpeed) #creates table object
prop.table(tb7, margin = 2) #crosstab with proportions
```

```{r}
#Looking at the relationship of RainTomorrow to AvgTemp
ggplot(rain_complete, aes(x= RainTomorrow, y= AvgTemp)) + geom_boxplot() 

#table view
tb8 = table(rain_complete$RainTomorrow, rain_complete$AvgTemp) #creates table object
prop.table(tb8, margin = 2) #crosstab with proportions
```

```{r}
#Looking at the relationship of RainTomorrow to MaxTemp
ggplot(rain_complete, aes(x= MaxTemp, fill= RainTomorrow)) + geom_bar() 

#table view
tb9 = table(rain_complete$RainTomorrow, rain_complete$MaxTemp) #creates table object
prop.table(tb9, margin = 2) #crosstab with proportions
```

```{r}
#Looking at the relationship of RainTomorrow to MinTemp
ggplot(rain_complete, aes(x= MinTemp, fill= RainTomorrow)) + geom_bar() 

#table view
tb10 = table(rain_complete$RainTomorrow, rain_complete$MinTemp) #creates table object
prop.table(tb10, margin = 2) #crosstab with proportions
```



#Finding Most Significant Variables using Backward Stepwise
```{r}

allmod =glm(RainTomorrow~., rain_complete, family = "binomial")
summary(allmod)

#backward
backmod =stepAIC(allmod, direction = "backward", trace = FALSE)
summary(backmod)

```

From this model, I have determined that RainToday, Rainfall, MaxTemp, MinTemp, WindGustSpeed, AvgHumidity, AvgPressure, AvgWindSpeed, and AvgTemp are significant indicators of if it will RainTomorrow. WindGustDirection is conditionally significant, in that only three directions [NNW, N, SSW] showed to have a significant impact on rain, with the N Wind having the most significant impact.


Lastly, before moving onto part two, I will split the data into my training and testing sets with a set seed of 1234.
###Creating Training and Testing Sets
```{r}
set.seed(1234) #sets random number seed for cross validation
train.rows = createDataPartition(y = rain$RainTomorrow, p=0.7, list= FALSE)
train = rain[train.rows,]
test = rain[-train.rows,]
```

